---
title: "helpAR (notebook version)"
author: "Cyrus Eosphoros"
date: 2022.04.29
output: html_notebook
---

*`helpAR`* uses `data.table` to generate unicode IPA substitutions for
ARPAbet. Its output is a vector in the original order.

Most recent benchmarks (2022.04.29) are \~30ms time elapsed and \~25MiB
memory usage for 350,000 rows of input.

[It lives here.](https://github.com/cyrusae/helpAR) Report issues [here](https://github.com/cyrusae/helpAR/issues).

Open in RStudio and follow along below to use.

## You will need data.table:

```{r}
if (!require(data.table)) { install.packages("data.table") }
library(data.table)
```

## Generate translation tables:

```{r arpa-to-ipa}
mARPA <- c("a", "@", "A", "c", "W", "x", "Y", "E", "R", "e", "I", "X", "i", "o", "O", "U", "u", "b", "C", "d", "D", "F", "L", "M", "N", "f", "g", "h", "J", "k", "l", "m", "n", "G", "p", "Q", "r", "s", "S", "t", "T", "v", "w", "H", "y", "z", "Z")
clARPA <- c("AA", "AE", "AH", "AO", "AW", "AX", "AY", "EH", "ER", "EY", "IH", "IX", "IY", "OW", "OY", "UH", "UW", "B", "CH", "D", "DH", "DX", "EL", "EM", "EN", "F", "G", "H", "JH", "K", "L", "M", "N", "NG", "P", "Q", "R", "S", "SH", "T", "TH", "V", "W", "WH", "J", "Z", "ZH")
lcARPA <- tolower(clARPA)
stress1 <- c("AA1", "AE1", "AH1", "AO1", "AW1", "AX1", "AY1", "EH1", "ER1", "EY1", "IH1", "IX1", "IY1", "OW1", "OY1", "UH1", "UW1")
stress1b <- tolower(stress1)
stress2 <- c("AA2", "AE2", "AH2", "AO2", "AW2", "AX2", "AY2", "EH2", "ER2", "EY2", "IH2", "IX2", "IY2", "OW2", "OY2", "UH2", "UW2")
stress2b <- tolower(stress2)
stress0 <- c("AA0", "AE0", "AH0", "AO0", "AW0", "AX0", "AY0", "EH0", "ER0", "EY0", "IH0", "IX0", "IY0", "OW0", "OY0", "UH0", "UW0")
stress0b <- tolower(stress0)
weirds <- c("AXR", "UX", "HH", "NX", "ENG")
loweirds <- tolower(weirds)
weIPA <- c("\u025A", "\u0289", "h", "\u027E\u0303", "\u014B\u030D")
unIPA <- c("\u0061", "\u00E6", "\u0250", "\u0254", "a\u028A", "\u0259", "a\u026A", "\u025B", "\u025d", "e\u026A", "\u026A", "\u0268", "i", "o\u028A", "\u0254\u026A", "\u028A", "u", "b", "\u02A7", "d", "\u00F0", "\u027E", "l\u0329", "m\u0329", "n\u0329", "f", "g", "h", "\u02A4", "k", "l", "m", "n", "\u014B", "p", "\u0294", "\u0279", "s", "\u0283", "t", "\u03B8", "v", "w", "\u028D", "y", "z", "\u0292")
vowIPA <- c("\u0061", "\u00E6", "\u0250", "\u0254", "a\u028A", "\u0259", "a\u026A", "\u025B", "\u025D", "e\u026A", "\u026A", "\u0268", "i", "o\u028A", "\u0254\u026A", "\u028A", "u")
#make the reference tables
arM <- data.table(
  ar = mARPA,
  ip = unIPA
)
setkey(arM, ar)
rm(mARPA)
arC <- data.table(
  ar = clARPA,
  ip = unIPA
)
setkey(arC, ar)
rm(clARPA)
arL <- data.table(
  ar = lcARPA,
  ip = unIPA
)
setkey(arL, ar)
rm(lcARPA, unIPA)
alt1 <- data.table(
  ar = stress1,
  ip = vowIPA
)
setkey(alt1, ar)
rm(stress1)
alt2 <- data.table(
  ar = stress1b,
  ip = vowIPA
)
setkey(alt2, ar)
rm(stress1b)
alt3 <- data.table(
  ar = stress2,
  ip = vowIPA
)
setkey(alt3, ar)
rm(stress2)
alt4 <- data.table(
  ar = stress2b,
  ip = vowIPA
)
setkey(alt4, ar)
rm(stress2b)
alt5 <- data.table(
  ar = stress0,
  ip = vowIPA
)
setkey(alt5, ar)
rm(stress0)
alt6 <- data.table(
  ar = stress0b,
  ip = vowIPA
)
setkey(alt6, ar)
rm(stress0b, vowIPA)
weird <- data.table(
  ar = weirds,
  ip = weIPA
)
setkey(weird, ar)
rm(weirds)
loweird <- data.table(
  ar = loweirds,
  ip = weIPA
)
setkey(loweird, ar)
rm(loweirds, weIPA)
```

Leave these in working memory for as long as you intend to use
*`helpAR`* to avoid repeating work.

## The helpAR functions:

```{r helpAR}
#turns input vector into data table
savAR <- function(vecc) {
  dt <- data.table(
    rw = vecc,
    ip = as.character(NA),
    or = 1:length(vecc)
  )
  setkey(dt, rw) 
  dt
}

#makes table of unique phones from input
parsAR <- function(un) {
  op <- data.table(
    rw = un,
    ip = as.character(NA)
  )
  setkey(op, rw)
  op
}

#the actual translation heavy lifting happens here
makAR <- function(dt) {
  dt <- arM[dt]
  dt[is.na(ip), ip := arM[.(.SD), ip]]
  dt[is.na(ip), ip := arC[.(.SD), ip]]
  dt[is.na(ip), ip := arL[.(.SD), ip]]
  dt[is.na(ip), ip := alt1[.(.SD), ip]]
  dt[is.na(ip), ip := alt2[.(.SD), ip]]
  dt[is.na(ip), ip := alt3[.(.SD), ip]]
  dt[is.na(ip), ip := alt4[.(.SD), ip]]
  dt[is.na(ip), ip := alt5[.(.SD), ip]]
  dt[is.na(ip), ip := alt6[.(.SD), ip]]
  dt[is.na(ip), ip := weird[.(.SD), ip]]
  dt[is.na(ip), ip := loweird[.(.SD), ip]]
  dt
}

#turn the ipa column of our working table into a vector and return
sendAR <- function(res) {
  setorder(res, or)
  vc <- res[[2]]
  vc
}

#replace anything that couldn't be identified as IPA with whatever the original input was
fixAR <- function(x) {
  qc <- data.table(
    ar = x,
    ip = x
  )
  setkey(qc, ar)
  qc
}

helpAR <- function(x) {
  un <- unique(x)
  qc <- fixAR(un)
  res <- savAR(x)
  y <- makAR(parsAR(un))
  y <- y[is.na(ip), ip := qc[.(.SD), ip]][, i.ip := NULL]
  y <- res[is.na(ip), ip := y[.(.SD), ip]]
  yeet <- sendAR(y)
  yeet
}
```

## How to use

1.  Open this notebook in the same RStudio session you're using to do
    work.

2.  Run the above chunks in order. You have now loaded the functions for
    use in your session.

3.  Get your output like this:

    `result <- helpAR([YOUR TABLE]$[PHONES COLUMN])`

`result` will be a vector with length and ordering equal to the input.
Replace the original phones column with it if you're feeling confident,
or add it as a new column to be cautious.

## Cleanup

Remove the translation tables:

```{r table-removing}
rm(arM, arC, arL, alt1, alt2, alt3, alt4, alt5, alt6, weird, loweird)
```

Clean the functions out of working memory:

```{r function-removing}
rm(savAR, parsAR, fixAR, sendAR, makAR, helpAR)
```

## Caveats

-   Requires that you feed it a vector in which every item is a *single*
    phoneme. Giving a column of a tibble, table, or data.frame in which
    each cell is a phone will work fine.
-   Currently, *`helpAR`* a) disregards stress and b) only recognizes
    digits 0-2 for stress. (Stress is being handled with the `alt_`
    tables, so changing that should be fairly transparent copy and paste
    if you need that in your life.)
-   Doesn't have full TIMIT support yet ("ENG" is supported, no others);
    you'd want to add it to `weird` if you need it.
-   I'm not a phonologist. I am, specifically, a linguist who avoids
    phonology as much as I am able because I have neurological issues
    that prevent me from recognizing IPA. I have proofread the lists
    helpAR uses multiple times and *believe* all the IPA is assigned to
    the right ARPAbet and vice versa. [Tell me if they're not.](https://github.com/cyrusae/helpAR/issues)

## Attribution

If you use *`helpAR`* in your own work, you should be transparent about
that for two reasons:

1.  Any script you hand someone else that tries to call on it without
    context will mysteriously fail, which is at best embarrassing.
2.  It's licensed under LGPL and you shouldn't be a dick.

## Future

This will be a package eventually. I need to work out whether a) there
are any glaring kinks that didn't come up in my initial tests and b) it
could be faster in C++ first. As of 2022.04.29 it works when I test it but (as you can see) there are not example tests in this document yet.